---
title: "Wei et al.: Gsdmd BBB"
subtitle: "Functional Analysis"
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'functional_analysis.html'))})
---

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
set.seed(42)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(clusterProfiler)
  library(org.Mm.eg.db)
  library(msigdbr)
})
```

```{r}
if(!dir.exists("../outs/functional_analysis/classification")) {dir.create("../outs/functional_analysis/classification")}
if(!dir.exists("../outs/functional_analysis/enrichment")) {dir.create("../outs/functional_analysis/ora")}
if(!dir.exists("../outs/functional_analysis/gsea")) {dir.create("../outs/functional_analysis/fcs")}
```

# Functional Analysis

## Load Data

```{r}
degs.cell_subtype.list <- readRDS("../outs/de_analysis/degs_cell_subtype_list.rds")
degs.cell_subtype.lps.wt_gdko.list <- readRDS("../outs/de_analysis/degs_cell_subtype_lps_wt_gdko_list.rds")
```

```{r}
degs.cell_type.list <- readRDS("../outs/de_analysis/degs_cell_type_list.rds")
degs.cell_type.lps.wt_gdko.list <- readRDS("../outs/de_analysis/degs_cell_type_lps_wt_gdko_list.rds")
```

```{r}
csvd.lps.mks.cell_subtype <- readRDS("../outs/de_analysis/csvd_lps_mks_cell_subtype.rds")
```

### 1 degs.cell_subtype.list

#### 1.1 Functional Classification

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_subtypes = names(degs.cell_subtype.list)
genotypes = c("WT", "GDKO")

ggo.cell_subtype.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_subtype.list[[x]][[g]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform functional grouping
          if (!is.null(degs_entrez)){
            return(
              groupGO(gene = degs_entrez,
                      OrgDb = org.Mm.eg.db,
                      keyType = "ENTREZID",
                      ont = o,
                      readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ggo.cell_subtype.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_subtype.list[[x]][[g]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform functional grouping
          if (!is.null(degs_entrez)){
            return(
              groupGO(gene = degs_entrez,
                      OrgDb = org.Mm.eg.db,
                      keyType = "ENTREZID",
                      ont = o,
                      readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ggo.cell_subtype.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ggo.cell_subtype.up.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ggo.cell_subtype.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ggo.cell_subtype.down.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save results.
```{r}
saveRDS(ggo.cell_subtype.up.list, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.down.list, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.up.dfs, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.down.dfs, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

#### 1.2 Functional Enrichment - Over Representation Analysis (ORA)

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_subtypes = names(degs.cell_subtype.list)
genotypes = c("WT", "GDKO")

background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_subtype.list[[1]][[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

ego.cell_subtype.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_subtype.list[[x]][[g]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_subtype.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_subtype.list[[x]][[g]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ego.cell_subtype.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ego.cell_subtype.up.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_subtype.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ego.cell_subtype.down.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save Results.
```{r}
saveRDS(ego.cell_subtype.up.list, paste0("../outs/functional_analysis/ora/ego_cell_subtype_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.down.list, paste0("../outs/functional_analysis/ora/ego_cell_subtype_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.up.dfs, paste0("../outs/functional_analysis/ora/ego_cell_subtype_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.down.dfs, paste0("../outs/functional_analysis/ora/ego_cell_subtype_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

### 2 degs.cell_subtype.lps.wt_gdko.list

#### 2.1 Functional Classification

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_subtypes = names(degs.cell_subtype.list)

ggo.cell_subtype.lps.wt_gdko.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        # # Show progress, for debug
        # print(x)
        
        # Parse DEGs list
        degs = degs.cell_subtype.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
        degs_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = degs,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(degs_entrez)){
          return(
            groupGO(gene = degs_entrez,
                    OrgDb = org.Mm.eg.db,
                    keyType = "ENTREZID",
                    ont = o,
                    readable = TRUE)
            )
          } else { return(NULL) }
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ggo.cell_subtype.lps.wt_gdko.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        # # Show progress, for debug
        # print(x)
        
        # Parse DEGs list
        degs = degs.cell_subtype.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
        degs_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = degs,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(degs_entrez)){
          return(
            groupGO(gene = degs_entrez,
                    OrgDb = org.Mm.eg.db,
                    keyType = "ENTREZID",
                    ont = o,
                    readable = TRUE)
            )
          } else { return(NULL) }
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ggo.cell_subtype.lps.wt_gdko.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        ggo.cell_subtype.lps.wt_gdko.up.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ggo.cell_subtype.lps.wt_gdko.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        ggo.cell_subtype.lps.wt_gdko.down.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save Results.
```{r}
saveRDS(ggo.cell_subtype.lps.wt_gdko.up.list, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_lps_wt_gdko_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.lps.wt_gdko.down.list, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_lps_wt_gdko_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.lps.wt_gdko.up.dfs, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_lps_wt_gdko_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ggo.cell_subtype.lps.wt_gdko.down.dfs, paste0("../outs/functional_analysis/classification/ggo_cell_subtype_lps_wt_gdko_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

#### 2.2 Functional Enrichment - Over Representation Analysis (ORA)

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_subtypes = names(degs.cell_subtype.list)  # "Ex-Neu.", "Inh-Neu.", "Oligo", "Astro", "OPCs", "Micro", "Endo"

background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_subtype.lps.wt_gdko.list[[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

ego.cell_subtype.lps.wt_gdko.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        # Parse DEGs list
        degs = degs.cell_subtype.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
        degs_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = degs,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(degs_entrez)){
          return(
            enrichGO(gene = degs_entrez,
                     OrgDb = org.Mm.eg.db,
                     keyType = "ENTREZID",
                     universe = background,
                     ont = o,
                     pAdjustMethod = "BH",
                     pvalueCutoff = 0.05,
                     qvalueCutoff = 0.2,
                     readable = TRUE)
            )} else { return(NULL) }
          },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_subtype.lps.wt_gdko.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        # Parse DEGs list
        degs = degs.cell_subtype.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
        degs_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = degs,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(degs_entrez)){
          return(
            enrichGO(gene = degs_entrez,
                     OrgDb = org.Mm.eg.db,
                     keyType = "ENTREZID",
                     universe = background,
                     ont = o,
                     pAdjustMethod = "BH",
                     pvalueCutoff = 0.05,
                     qvalueCutoff = 0.2,
                     readable = TRUE)
            )} else { return(NULL) }
          },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ego.cell_subtype.lps.wt_gdko.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        ego.cell_subtype.lps.wt_gdko.up.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_subtype.lps.wt_gdko.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_subtypes, 
      FUN = function(x){
        ego.cell_subtype.lps.wt_gdko.down.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save results.
```{r}
saveRDS(ego.cell_subtype.lps.wt_gdko.up.list,
        paste0("../outs/functional_analysis/ora/ego_cell_subtype_lps_wt_gdko_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.lps.wt_gdko.down.list,
        paste0("../outs/functional_analysis/ora/ego_cell_subtype_lps_wt_gdko_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.lps.wt_gdko.up.dfs, 
        paste0("../outs/functional_analysis/ora/ego_cell_subtype_lps_wt_gdko_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_subtype.lps.wt_gdko.down.dfs,
        paste0("../outs/functional_analysis/ora/ego_cell_subtype_lps_wt_gdko_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

### 3 degs.cell_type.list

#### 3.1 Functional Enrichment - Over Representation Analysis (ORA)

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_types = names(degs.cell_type.list)
genotypes = c("WT", "GDKO")

background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_type.list[[1]][[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

ego.cell_type.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_type.list[[x]][[g]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_type.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
          # Parse DEGs list
          degs = degs.cell_type.list[[x]][[g]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ego.cell_type.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ego.cell_type.up.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_type.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
        sapply(X = genotypes, FUN = function(g){
            ego.cell_type.down.list[[o]][[x]][[g]] |> as.data.frame()
          },
          simplify = FALSE, 
          USE.NAMES = TRUE
          )},
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save Results.
```{r}
saveRDS(ego.cell_type.up.list, paste0("../outs/functional_analysis/ora/ego_cell_type_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.down.list, paste0("../outs/functional_analysis/ora/ego_cell_type_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.up.dfs, paste0("../outs/functional_analysis/ora/ego_cell_type_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.down.dfs, paste0("../outs/functional_analysis/ora/ego_cell_type_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

#### 3.2 Functional Class Scoring (FCS) - Gene Set Enrichment Analysis (GSEA)

Neuronal signatures are weak. Try to recover them with GSEA.
```{r}
degs.cell_type.list.entrez <- lapply(
  X = degs.cell_type.list, 
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(g){
        return(
          g |>
            dplyr::mutate(entrez_id = mapIds(x = org.Mm.eg.db,
                                             keys = gene,
                                             keytype = "SYMBOL",
                                             column = "ENTREZID",
                                             multiVals = "first")) |>
            na.omit()
        )
        })
    })

cell_type.gene.list <- lapply(
  X = degs.cell_type.list.entrez, 
  FUN = function(ct){
    lapply(
      X = ct, 
      FUN = function(g){
        genelist = g[["avg_log2FC"]]
        names(genelist) = g[["entrez_id"]]
        genelist = sort(genelist, decreasing = TRUE)
        return(genelist)
        })
})
```

```{r}
m_t2g_C5_BP <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") |>
  dplyr::select(gs_name, entrez_gene)

cell_type.gsea.gobp.list <- lapply(
  X = cell_type.gene.list, 
  FUN = function(ct){
    lapply(X = ct, FUN = function(g){
      gsea = GSEA(g, TERM2GENE = m_t2g_C5_BP, eps = 0)
      gsea = setReadable(gsea, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
      return(gsea)
    })
})
```

```{r}
# C2: Curated Set, Canonical Pathways
m_t2g_C2 <- msigdbr(species = "Mus musculus", category = "C2") |>
    dplyr::select(gs_name, entrez_gene)

cell_type.gsea.C2.list <- lapply(
  X = cell_type.gene.list, 
  FUN = function(ct){
    lapply(X = ct, FUN = function(g){
      gsea = GSEA(g, TERM2GENE = m_t2g_C2, eps = 0)
      gsea = setReadable(gsea, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
      return(gsea)
    })
})
```

### 4 degs.cell_type.lps.wt_gdko.list

#### 4.1 Functional Enrichment

Setup test parameters.
```{r}
ABS_LFC_cutoff = 1
P_ADJ_cutoff = 0.01
```

```{r}
ont = c("BP", "CC", "MF")
cell_types = names(degs.cell_type.lps.wt_gdko.list)

background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_type.lps.wt_gdko.list[[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

ego.cell_type.lps.wt_gdko.up.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
          # Parse DEGs list
          degs = degs.cell_type.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC > ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_type.lps.wt_gdko.down.list <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
          # Parse DEGs list
          degs = degs.cell_type.lps.wt_gdko.list[[x]] |> dplyr::filter(avg_log2FC < -ABS_LFC_cutoff & p_val_adj < P_ADJ_cutoff) |> pull("gene")
          degs_entrez = tryCatch(
            mapIds(x = org.Mm.eg.db,
                   keys = degs,
                   column = "ENTREZID",
                   keytype = "SYMBOL",
                   multiVals = "first") |> na.omit(),
            error = function(e){ return(NULL) }
          )
          
          # Perform enrichment tests
          if (!is.null(degs_entrez)){
            return(
              enrichGO(gene = degs_entrez,
                       OrgDb = org.Mm.eg.db,
                       keyType = "ENTREZID",
                       universe = background,
                       ont = o,
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
              )
            } else { return(NULL) }
          },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Convert to dataframe for viewer's convenience.
```{r}
ego.cell_type.lps.wt_gdko.up.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
            ego.cell_type.lps.wt_gdko.up.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)

ego.cell_type.lps.wt_gdko.down.dfs <- sapply(
  X = ont,
  FUN = function(o){
    sapply(
      X = cell_types, 
      FUN = function(x){
            ego.cell_type.lps.wt_gdko.down.list[[o]][[x]] |> as.data.frame()
        },
      simplify = FALSE, 
      USE.NAMES = TRUE
    )},
    simplify = FALSE, 
    USE.NAMES = TRUE
)
```

Save Results.
```{r}
saveRDS(ego.cell_type.lps.wt_gdko.up.list, paste0("../outs/functional_analysis/ora/ego_cell_type_lps_wt_gdko_up_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.lps.wt_gdko.down.list, paste0("../outs/functional_analysis/ora/ego_cell_type_lps_wt_gdko_down_list_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.lps.wt_gdko.up.dfs, paste0("../outs/functional_analysis/ora/ego_cell_type_lps_wt_gdko_up_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
saveRDS(ego.cell_type.lps.wt_gdko.down.dfs, paste0("../outs/functional_analysis/ora/ego_cell_type_lps_wt_gdko_down_dfs_LFC_",ABS_LFC_cutoff,"_PADJ_",P_ADJ_cutoff,".rds"))
```

#### 4.2 Functional Class Scoring (FCS) - Gene Set Enrichment Analysis (GSEA)

```{r}
genelist.cell_type.lps.wt_gdko <- lapply(
  X = degs.cell_type.lps.wt_gdko.list, 
  FUN = function(ct){
    df = ct |>
      dplyr::mutate(entrez_id = mapIds(x = org.Mm.eg.db,
                                       keys = gene,
                                       keytype = "SYMBOL",
                                       column = "ENTREZID",
                                       multiVals = "first")) |>
      na.omit()
    genelist = df[["avg_log2FC"]]
    names(genelist) = df[["entrez_id"]]
    genelist = sort(genelist, decreasing = TRUE)
      return(genelist)
})
```

```{r}
# C5: GOBP ontology set
m_t2g_C5_BP <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") |>
  dplyr::select(gs_name, entrez_gene)

cell_type.lps.wt_gdko.gsea.C5.list <- lapply(
  X = genelist.cell_type.lps.wt_gdko, 
  FUN = function(ct){
    gsea = GSEA(ct, TERM2GENE = m_t2g_C5_BP, eps = 0)
    gsea = setReadable(gsea, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
    return(gsea)
    })
```

```{r}
# C2: Curated Set, Canonical Pathways
m_t2g_C2 <- msigdbr(species = "Mus musculus", category = "C2") |>
    dplyr::select(gs_name, entrez_gene)

cell_type.lps.wt_gdko.gsea.C2.list <- lapply(
  X = genelist.cell_type.lps.wt_gdko, 
  FUN = function(ct){
    gsea = GSEA(ct, TERM2GENE = m_t2g_C2, eps = 0)
    gsea = setReadable(gsea, OrgDb = org.Mm.eg.db, keyType = "ENTREZID")
    return(gsea)
    })
```

Save results.
```{r}
saveRDS(cell_type.lps.wt_gdko.gsea.C5.list,
        paste0("../outs/functional_analysis/fcs/cell_type_lps_wt_gdko_gsea_C5_list.rds"))
saveRDS(cell_type.lps.wt_gdko.gsea.C2.list,
        paste0("../outs/functional_analysis/fcs/cell_type_lps_wt_gdko_gsea_C2_list.rds"))
```

### 5 csvd.lps.mks.cell_subtype

#### 5.1 Functional Enrichment

```{r}
background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_subtype.lps.wt_gdko.list[[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

ego.bp.signature_genes.cell_subtype.list <- lapply(
  X = signature_genes.cell_subtype.list,
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(g){
        g_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = g,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(g_entrez)){
          return(
            enrichGO(gene = g_entrez,
                     OrgDb = org.Mm.eg.db,
                     keyType = "ENTREZID",
                     universe = background,
                     ont = "BP",
                     pAdjustMethod = "BH",
                     pvalueCutoff = 0.1,
                     # qvalueCutoff = 0.2,
                     readable = TRUE)
            )} else { return(NULL) }
      }
    )
  }
)
```

```{r}
ego.bp.signature_genes.cell_subtype.dfs <- lapply(
  X = ego.bp.signature_genes.cell_subtype.list,
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(ego){
        ego |> as.data.frame()
      }
    )
  }
)
```

```{r}
saveRDS(ego.bp.signature_genes.cell_subtype.list, "../outs/functional_analysis/ora/ego_bp_signature_genes_cell_subtype_list.rds")
saveRDS(ego.bp.signature_genes.cell_subtype.dfs, "../outs/functional_analysis/ora/ego_bp_signature_genes_cell_subtype_dfs.rds")
```

#### 5.2 KEGG Pathway ORA

```{r}
background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_subtype.lps.wt_gdko.list[[1]][["gene"]], 
  column = "ENTREZID", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

kk.entrez.signature_genes.cell_subtype.list <- lapply(
  X = signature_genes.cell_subtype.list,
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(g){
        g_entrez = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = g,
                 column = "ENTREZID",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(g_entrez)){
          return(
            enrichKEGG(
              gene = g_entrez,
              organism = "mmu",
              keyType = "kegg",
              pvalueCutoff = 0.05,
              pAdjustMethod = "BH",
              universe = background
              )
            )} else { return(NULL) }
      }
    )
  }
)
```

```{r}
kk.entrez.signature_genes.cell_subtype.dfs <- lapply(
  X = kk.entrez.signature_genes.cell_subtype.list,
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(ego){
        ego |> as.data.frame()
      }
    )
  }
)
```

```{r}
saveRDS(kk.entrez.signature_genes.cell_subtype.list, "../outs/functional_analysis/ora/kk_entrez_signature_genes_cell_subtype_list.rds")
```



```{r}
background = mapIds(
  x = org.Mm.eg.db, 
  keys = degs.cell_subtype.lps.wt_gdko.list[[1]][["gene"]], 
  column = "UNIPROT", 
  keytype = "SYMBOL", 
  multiVals = "first"
) %>% na.omit()

kk.uniprot.signature_genes.cell_subtype.list <- lapply(
  X = signature_genes.cell_subtype.list,
  FUN = function(ct){
    lapply(
      X = ct,
      FUN = function(g){
        g_uniprot = tryCatch(
          mapIds(x = org.Mm.eg.db,
                 keys = g,
                 column = "UNIPROT",
                 keytype = "SYMBOL",
                 multiVals = "first") |> na.omit(),
          error = function(e){ return(NULL) }
          )
        
        # Perform enrichment tests
        if (!is.null(g_uniprot)){
          return(
            enrichKEGG(
              gene = g_uniprot,
              organism = "mmu",
              keyType = "uniprot",
              pvalueCutoff = 0.05,
              pAdjustMethod = "BH",
              universe = background
              )
            )} else { return(NULL) }
      }
    )
  }
)
```