---
title: "Wei et al.: Gsdmd BBB"
subtitle: 'Preprocessing workflow for GDKO-LPS'
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'gdko_lps_mo_preprocessing.html'))})
---

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
WRITE_CACHE = FALSE  ## Set to TRUE on first run
READ_CACHE = !WRITE_CACHE
```

```{r}
suppressPackageStartupMessages({
  library(SoupX)
  library(Seurat)
  library(DoubletFinder)
  library(scater)
})
```

```{r}
reticulate::use_python("/home/luolab/miniconda3/envs/r-reticulate/bin/python")
```

```{r}
if(!dir.exists("../outs/preprocessing")) {dir.create("../outs/preprocessing")}
```

For knitr to work, steps involving manual operations (i.e. `CellSelector()`) are omitted. Ids for manually selected cells were stashed to and loaded from the `.rda` file located in the `../miscs/` directory.

```{r}
load("../miscs/gdko_lps_ids.rda")
```

```{r}
set.seed(42)
```

# Pre-processing Workflow

## Decontamination

Using `SoupX` for decontamination. Before constructing the soupchannel, precompute Louvain clusters and umap embeddings which will be later used.

```{r}
filtered_counts.gdko_lps <- Read10X("../data/GDKO-LPS-MO-0614_reanalysis/outs/filtered_feature_bc_matrix/")
cells <- CreateSeuratObject(counts = filtered_counts.gdko_lps, project = "GD_BBB")
# Standard Seurat Workflow. For conciseness, future runs of the workflow will use verbose = FALSE
cells <- cells |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA() |>
  FindNeighbors(dims = 1:50) |>
  FindClusters() |>
  RunUMAP(dims = 1:50)
```

```{r}
# Load data and estimate soup profile
sc = load10X("../data/GDKO-LPS-MO-0614_reanalysis/outs/")
# Set clusters and DR
sc = setClusters(sc, cells$RNA_snn_res.0.8)
sc = setDR(sc, Embeddings(object = cells, reduction = "umap")[colnames(sc$toc), c("UMAP_1","UMAP_2")])
# Estimate rho
sc = autoEstCont(sc)
# Clean the data
out = adjustCounts(sc, roundToInt = T)  # stochastic, don't forget to set.seed() for reproducibility
```

```{r}
if(WRITE_CACHE){
  DropletUtils::write10xCounts(path = "../data/GDKO-LPS-MO-0614_reanalysis/outs/strainedCounts.h5", x = out, type = "HDF5", overwrite = TRUE)  ## Save corrected matrix
}
```

```{r}
if(READ_CACHE){out <- Read10X_h5("../data/GDKO-LPS-MO-0614_reanalysis/outs/strainedCounts.h5")}

cells.decont <- CreateSeuratObject(counts = out, project = "GD_BBB")
cells.decont <- cells.decont |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Several cluster (6,15, subset of 11) shows very low level of recovered genes (nFeature_RNA) and high level of mitochondrial RNA species. These are characteristics of empty droplets, but CellRanger have mistakenly recognized them as cells. Here, we filter them out manually.

```{r fig.width=9, fig.height=6}
((DimPlot(cells.decont, group.by = "seurat_clusters", label = T) + NoLegend()) | DimPlot(cells.decont, cells.highlight = WhichCells(cells.decont, idents = c(6,15))) + ggtitle("Cells to Remove\n(Putative Empty droplets)") + NoLegend())/
  (FeaturePlot(cells.decont, features = c("nFeature_RNA","mt-Nd3")))
```

```{r}
# emptydrop_clusters1_ids <- c()
# emptydrop_clusters1_ids <- union(emptydrop_clusters1_ids, CellSelector(FeaturePlot(cells.decont, features = c("nFeature_RNA"))))
DimPlot(cells.decont, cells.highlight = emptydrop_clusters1_ids)
```

```{r}
edc1 <- subset(cells.decont, cells = emptydrop_clusters1_ids)
edc1 <- edc1 |>
  NormalizeData(verbose=FALSE) |>
  FindVariableFeatures(verbose=FALSE) |>
  ScaleData(verbose=FALSE) |>
  RunPCA(verbose=FALSE) |>
  FindNeighbors(dims = 1:40, verbose=FALSE) |>
  FindClusters(algorithm=4, verbose=FALSE) |>
  RunUMAP(dims = 1:40, verbose=FALSE)
```
```{r}
((DimPlot(edc1, group.by = "seurat_clusters", label = T) + NoLegend()) | DimPlot(edc1, cells.highlight = WhichCells(edc1, idents = c(3,10,12))) + ggtitle("Cells to Remove\n(Putative Empty droplets)") + NoLegend())/
  (FeaturePlot(edc1, features = c("nFeature_RNA","mt-Nd3")))
```

```{r}
# emptydrop_clusters2_ids <- CellSelector(FeaturePlot(cells.decont, features = c("nFeature_RNA")))
DimPlot(cells.decont, cells.highlight = emptydrop_clusters2_ids)
```

```{r}
edc2 <- subset(cells.decont, cells = emptydrop_clusters2_ids)
edc2 <- edc2 |>
  NormalizeData(verbose=FALSE) |>
  FindVariableFeatures(verbose=FALSE) |>
  ScaleData(verbose=FALSE) |>
  RunPCA(verbose=FALSE) |>
  FindNeighbors(dims = 1:30, verbose=FALSE) |>
  FindClusters(algorithm=4, verbose=FALSE) |>
  RunUMAP(dims = 1:30, verbose=FALSE)
```

```{r}
((DimPlot(edc2, group.by = "seurat_clusters", label = T) + NoLegend()) | DimPlot(edc2, cells.highlight = WhichCells(edc2, idents = c(1,5))) + ggtitle("Cells to Remove\n(Putative Empty droplets)") + NoLegend())/
  (FeaturePlot(edc2, features = c("nFeature_RNA","mt-Nd3")))
```

```{r}
empty_drop_ids <- c(WhichCells(edc1, expression = seurat_clusters %in% c(3,10,12)), WhichCells(edc2, expression = seurat_clusters %in% c(1,5)))
cells.clean <- subset(
  cells.decont,
  cells = empty_drop_ids,
  invert = TRUE
)
```

## Doublet Removal

We use `DoubletFinder` to remove doublets. We aimed at ~8000 cell recovery during wet experiment. According to 10X, this corresponds to ~0.06 doublet rates, thus we set `percent_est = 0.06`.

```{r}
PCs = 1:10
percent_est = 0.06
pN = 0.25
```

```{r}
cells.clean <- cells.clean |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

```{r}
suppressed_msg <- capture.output({
  sweep.res.list_cells <- paramSweep_v3(cells.clean, PCs = PCs, sct = FALSE)
  sweep.stats_cell <- summarizeSweep(sweep.res.list_cells, GT = FALSE)
  bcmvn_cell <- find.pK(sweep.stats_cell)
  pk_cell <- as.numeric(levels(bcmvn_cell[, 2]))[which.max(bcmvn_cell[,5])]
  homotypic.prop <- modelHomotypic(annotations <- cells.clean@meta.data$seurat_clusters)
  nExp_poi <- round(percent_est*nrow(cells.clean@meta.data))
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  cells.clean <- doubletFinder_v3(cells.clean, PCs, pN, pK = pk_cell, nExp = nExp_poi.adj, reuse.pANN = FALSE, sct = FALSE)
  cells.clean$doubletfinder <- cells.clean@meta.data[grep("DF.",names(cells.clean@meta.data))]
  # cells.filtered <- subset(cells.clean, subset = doubletfinder != "Doublet")
})
```

Beware, sometimes we may encounter over removal! As a visual check, inspect the predicted doublets and retain critical populations.

```{r fig.width=9, fig.height=3.5}
p1 <- DimPlot(cells.clean, cells.highlight = WhichCells(cells.clean, expression = doubletfinder == "Doublet")) +
  scale_color_manual(labels = c("NA","Doublets"), values = c("gray","#CC0033")) +
  ggtitle("Predicted Doublets") + NoAxes()
p2 <- DimPlot(cells.clean, group.by = "seurat_clusters", label = T, pt.size = .5) + NoLegend() + NoAxes()
p1 | p2
```

```{r}
# correct_singlet_ids <- c()
# correct_singlet_ids <- union(correct_singlet_ids, CellSelector(p1))
# correct_doublet_ids <- CellSelector(p1)
Idents(cells.clean) <- "doubletfinder"
cells.clean <- SetIdent(cells.clean, cells = correct_singlet_ids, value = "Singlet")
cells.clean <- SetIdent(cells.clean, cells = correct_doublet_ids, value = "Doublet")
cells.clean$doubletfinder_fixed <- Idents(cells.clean)
cells.filtered <- subset(cells.clean, subset = doubletfinder_fixed != "Doublet")
```

## Quality Control

Using `scater` for QC.

```{r}
Seurat_Obj_Diet <- DietSeurat(cells.filtered)
cells.sce <- as.SingleCellExperiment(Seurat_Obj_Diet)

## Cell level QC
per.cell <- perCellQCMetrics(cells.sce, subsets = list(Mito = grep("^mt-", rownames(cells.sce)),
                                                       Ribo = grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))))
qc.stats <- quickPerCellQC(per.cell, percent_subsets=c("subsets_Mito_percent"))
colSums(as.matrix(qc.stats))

colData(cells.sce) <- cbind(colData(cells.sce), cbind(per.cell, qc.stats))
cells.sce.filtered <- cells.sce[,!cells.sce$discard]

## Feature level QC
keep_feature <- nexprs(cells.sce.filtered, byrow=TRUE) > 0
## Remove mitochondrial and ribosomal genes which does not contribute to analysis (i.e. cell clustering)
keep_feature[grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))] <- FALSE
keep_feature[grep("mt-", rownames(cells.sce))] <- FALSE
cells.sce.filtered <- cells.sce.filtered[keep_feature,]
dim(cells.sce.filtered)

gdko_lps <- as.Seurat(cells.sce.filtered)
gdko_lps$ident <- NULL
```

```{r}
gdko_lps <- gdko_lps |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Summary plot of some canonical marker genes for typical brain cell types.

```{r}
obj <- gdko_lps

f_to_plot <- c("nFeature_RNA","Cldn5","Kcnj8","Aqp4","Mbp","Pdgfra",
               "Slc17a7","Slc17a6","Gad2","Cx3cr1","Col1a1",
               "Gsdmd","Casp4","Cd14")
obj <- ScaleData(obj, features = c(VariableFeatures(obj), f_to_plot))
f_list <- FeaturePlot(obj, features = f_to_plot, reduction = "umap", combine = F)
for (i in 1:length(f_list)){
  f_list[[i]] <- f_list[[i]] + 
    NoLegend() +
    NoAxes()
}
f_list <- append(x = f_list, values = list((DimPlot(obj, label = T) + NoLegend() + NoAxes())), after = 0)
```

```{r fig.width=12, fig.height=9}
g <- ggpubr::ggarrange(plotlist = f_list)
g
```

```{r}
ggsave(
  filename = "gdko_lps_preprocessed_flist.png",
  plot = g,
  device = "png",
  path = "../outs/preprocessing",
  width = 12, height = 10,
  bg = "white"
)
```

## Cell Type Annotation

### Semi-supervised classification

To begin with, we perform an initial round of cell type annotation on the preprocessed, "stand-alone" (e.g. non-integrated) object. These labels will be cross-referenced later in the integrated dataset to check the accuracy of data integration.

```{r}
Idents(gdko_lps) <- "seurat_clusters"
gdko_lps <- RenameIdents(
  gdko_lps, 
  `4` = "Astro", `23` = "Astro", 
  `1` = "Ex-Neu.", `2` = "Ex-Neu.", `3` = "Ex-Neu.", `5` = "Ex-Neu.", `6` = "Ex-Neu.", `7` = "Ex-Neu.", `10` = "Ex-Neu.", `11` = "Ex-Neu.", `15` = "Ex-Neu.", `17` = "Ex-Neu.", `25` = "Ex-Neu.",
  `12` = "Inh-Neu.", `14` = "Inh-Neu.", `16` = "Inh-Neu.", `22` = "Inh-Neu.", 
  `13` = "Micro",
  `19` = "BVs", `20` = "BVs",
  `8` = "OPCs", `24` = "OPCs",
  `9` = "Oligo", `18` = "Oligo", `21` = "Oligo"
  )
gdko_lps$cell_type_leiden <- Idents(gdko_lps)
```

```{r fig.width=12, fig.height=5}
d_list <- lapply(levels(gdko_lps$cell_type_leiden), function(x){
  DimPlot(gdko_lps, cells.highlight = WhichCells(gdko_lps, expression = cell_type_leiden == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
ggpubr::ggarrange(plotlist = d_list, ncol = 4, nrow = 2)
```

### Manual curation

Leiden clustering does most of the job just fine.

```{r}
Idents(gdko_lps) <- "cell_type_leiden"
```

Subsetting out Blood Vessel (BV) population to refine their labels. 

```{r fig.width=10, fig.height=5}
bv <- subset(gdko_lps, cells = WhichCells(gdko_lps, expression = cell_type_leiden == "BVs"))
bv <- bv |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:20, verbose = FALSE)
```

```{r fig.width=12, fig.height=5}
f_list <- FeaturePlot(bv, features = c("Pecam1","Cldn5","Kcnj8","Col1a1","Mbp","Acta2","Map2","Slc17a7"), combine = FALSE)
ggpubr::ggarrange(plotlist = f_list, ncol = 4, nrow = 2)
```

```{r}
## Skip the manual steps and use stashed ids
# bv_endo_ids <- CellSelector(FeaturePlot(bv, features = "Pecam1"))
# bv_peri_ids <- c()
# bv_peri_ids <- union(bv_peri_ids, CellSelector(FeaturePlot(bv, features = "Kcnj8")))
# bv_vlmc_ids <- CellSelector(FeaturePlot(bv, features = "Col1a1"))
# bv_oligo_ids <- CellSelector(FeaturePlot(bv, features = "Mbp"))
# bv_exneu_ids <- CellSelector(FeaturePlot(bv, features = "Slc17a7"))
bv <- SetIdent(bv, cells = bv_endo_ids, value = "Endo")
bv <- SetIdent(bv, cells = bv_peri_ids, value = "Peri")
bv <- SetIdent(bv, cells = bv_vlmc_ids, value = "VLMCs")
bv <- SetIdent(bv, cells = bv_oligo_ids, value = "Oligo")
bv <- SetIdent(bv, cells = bv_exneu_ids, value = "Ex-Neu.")
bv$cell_type_curated <- Idents(bv)
```

```{r fig.width=9, fig.height=4}
d_list <- lapply(levels(bv$cell_type_curated), function(x){
  DimPlot(bv, cells.highlight = WhichCells(bv, expression = cell_type_curated == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
d_list <- append(x = d_list, values = list((DimPlot(bv) + NoAxes() + NoLegend())), after = 0)
ggpubr::ggarrange(plotlist = d_list, ncol = 3, nrow = 2)
```

```{r}
# Update labels in master object
for(i in levels(bv$cell_type_curated)){
  gdko_lps <- SetIdent(gdko_lps, cells = WhichCells(bv, expression = cell_type_curated == i), value = i)
}
```

Refine labels in Monocyte cluster.

```{r fig.width=10, fig.height=5}
# mono_ids <- CellSelector(DimPlot(gdko_lps))
mono <- subset(gdko_lps, cells = mono_ids)
mono <- mono |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:20, verbose = FALSE)
```

```{r}
# mono_micro_ids <- CellSelector(FeaturePlot(mono, features = "Cx3cr1"))
# mono_oligo_ids <- CellSelector(FeaturePlot(mono, features = "Mbp"))
# mono_astro_ids <- c()
# mono_astro_ids <- union(mono_astro_ids, CellSelector(FeaturePlot(mono, features = "Aqp4")))
# mono_opcs_ids <- c()
# mono_opcs_ids <- union(mono_opcs_ids, CellSelector(FeaturePlot(mono, features = "Pdgfra")))
# mono_exneu_ids <- CellSelector(FeaturePlot(mono, features = "Slc17a7"))
# mono_inhneu_ids <- CellSelector(FeaturePlot(mono, features = "Gad2"))
mono <- SetIdent(mono, cells = mono_micro_ids, value = "Micro")
mono <- SetIdent(mono, cells = mono_oligo_ids, value = "Oligo")
mono <- SetIdent(mono, cells = mono_astro_ids, value = "Astro")
mono <- SetIdent(mono, cells = mono_opcs_ids, value = "OPCs")
mono <- SetIdent(mono, cells = mono_exneu_ids, value = "Ex-Neu.")
mono <- SetIdent(mono, cells = mono_inhneu_ids, value = "Inh-Neu.")
mono$cell_type_curated <- Idents(mono)
```

```{r fig.width=10, fig.height=5}
d_list <- lapply(levels(mono$cell_type_curated), function(x){
  DimPlot(mono, cells.highlight = WhichCells(mono, expression = cell_type_curated == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
ggpubr::ggarrange(plotlist = d_list, ncol = 3, nrow = 2)
```

```{r}
# Update labels in master object
for(i in levels(mono$cell_type_curated)){
  gdko_lps <- SetIdent(gdko_lps, cells = WhichCells(mono, expression = cell_type_curated == i), value = i)
}
```

```{r}
# Stash curated labels
gdko_lps$cell_type_standalone <- Idents(gdko_lps)

# Append sample info
gdko_lps$sample = "GDKO LPS"
gdko_lps$genotype = "GDKO"
gdko_lps$treatment = "LPS"
```

## Save Results

```{r}
if(WRITE_CACHE){
  save(emptydrop_clusters1_ids, emptydrop_clusters2_ids, empty_drop_ids, correct_doublet_ids, correct_singlet_ids, bv_endo_ids, bv_exneu_ids, bv_oligo_ids, bv_peri_ids,
       bv_vlmc_ids, mono_ids, mono_exneu_ids, mono_inhneu_ids, mono_astro_ids, mono_micro_ids, mono_oligo_ids, mono_opcs_ids, 
       file = "../miscs/gdko_lps_ids.rda")
  SeuratDisk::SaveH5Seurat(object = gdko_lps, filename = "../data/GDKO-LPS-MO-0614_reanalysis/gdko_lps_preprocessed.h5Seurat", overwrite = T, verbose = F)
}
```

# Session info

```{r}
sessionInfo()
```
