---
title: "Wei et al.: Gsdmd BBB"
subtitle: 'Preprocessing workflow for GDKO-PBS'
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'gdko_pbs_mo_preprocessing.html'))})
---

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
WRITE_CACHE = FALSE  ## Set to TRUE on first run
READ_CACHE = !WRITE_CACHE
```

```{r}
suppressPackageStartupMessages({
  library(SoupX)
  library(Seurat)
  library(DoubletFinder)
  library(scater)
})
```

```{r}
reticulate::use_python("/home/luolab/miniconda3/envs/r-reticulate/bin/python")
```

```{r}
if(!dir.exists("../outs/preprocessing")) {dir.create("../outs/preprocessing")}
```

For knitr to work, steps involving manual operations (i.e. `CellSelector()`) are omitted. Ids for manually selected cells were stashed to and loaded from the `.rda` file located in the `../miscs/` directory.

```{r}
load("../miscs/gdko_pbs_ids.rda")
```

```{r}
set.seed(42)
```

# Pre-processing Workflow

## Decontamination

Using `SoupX` for decontamination. Before constructing the soupchannel, precompute Louvain clusters and umap embeddings which will be later used.

```{r}
filtered_counts.gdko_pbs <- Read10X("../data/GDKO-PBS-MO-0614_reanalysis/outs/filtered_feature_bc_matrix/")
cells <- CreateSeuratObject(counts = filtered_counts.gdko_pbs, project = "GD_BBB")
# Standard Seurat Workflow. For conciseness, future runs of the workflow will use verbose = FALSE
cells <- cells |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA() |>
  FindNeighbors(dims = 1:50) |>
  FindClusters() |>
  RunUMAP(dims = 1:50)
```

```{r}
# Load data and estimate soup profile
sc = load10X("../data/GDKO-PBS-MO-0614_reanalysis/outs/")
# Set clusters and DR
sc = setClusters(sc, cells$RNA_snn_res.0.8)
sc = setDR(sc, Embeddings(object = cells, reduction = "umap")[colnames(sc$toc), c("UMAP_1","UMAP_2")])
# Estimate rho
sc = autoEstCont(sc)
# Clean the data
out = adjustCounts(sc, roundToInt = T)  # stochastic, don't forget to set.seed() for reproducibility
```

```{r}
if(WRITE_CACHE){
  DropletUtils::write10xCounts(path = "../data/GDKO-PBS-MO-0614_reanalysis/outs/strainedCounts.h5", x = out, type = "HDF5", overwrite = TRUE)  ## Save corrected matrix
}
```

```{r}
if(READ_CACHE){out <- Read10X_h5("../data/GDKO-PBS-MO-0614_reanalysis/outs/strainedCounts.h5")}

cells.decont <- CreateSeuratObject(counts = out, project = "GD_BBB")
cells.decont <- cells.decont |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Several cluster (3,4,16) shows very low level of recovered genes (nFeature_RNA) and high level of mitochondrial RNA species. These are characteristics of empty droplets, but CellRanger have mistakenly recognized them as cells. Here, we filter them out manually.

```{r fig.width=9, fig.height=6}
((DimPlot(cells.decont, group.by = "seurat_clusters", label = T) + NoLegend()) | DimPlot(cells.decont, cells.highlight = WhichCells(cells.decont, idents = c(3,4,16))) + ggtitle("Cells to Remove\n(Putative Empty droplets)") + NoLegend())/
  (FeaturePlot(cells.decont, features = c("nFeature_RNA","mt-Nd3")))
```

```{r}
cells.clean <- subset(
  cells.decont,
  idents = c(3,4,16),
  invert = TRUE
)
```

## Doublet Removal

We use `DoubletFinder` to remove doublets. We aimed at ~8000 cell recovery during wet experiment. According to 10X, this corresponds to ~0.06 doublet rates, thus we set `percent_est = 0.06`.

```{r}
PCs = 1:10
percent_est = 0.06
pN = 0.25
```

```{r}
cells.clean <- cells.clean |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

```{r}
suppressed_msg <- capture.output({
  sweep.res.list_cells <- paramSweep_v3(cells.clean, PCs = PCs, sct = FALSE)
  sweep.stats_cell <- summarizeSweep(sweep.res.list_cells, GT = FALSE)
  bcmvn_cell <- find.pK(sweep.stats_cell)
  pk_cell <- as.numeric(levels(bcmvn_cell[, 2]))[which.max(bcmvn_cell[,5])]
  homotypic.prop <- modelHomotypic(annotations <- cells.clean@meta.data$seurat_clusters)
  nExp_poi <- round(percent_est*nrow(cells.clean@meta.data))
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  cells.clean <- doubletFinder_v3(cells.clean, PCs, pN, pK = pk_cell, nExp = nExp_poi.adj, reuse.pANN = FALSE, sct = FALSE)
  cells.clean$doubletfinder <- cells.clean@meta.data[grep("DF.",names(cells.clean@meta.data))]
  # cells.filtered <- subset(cells.clean, subset = doubletfinder != "Doublet")
})
```

Beware, sometimes we may encounter over removal! As a visual check, inspect the predicted doublets. We notice that two clusters corresponding to Microglia/Macrophage (12) and Pericytes (17) contain many cells predicted as doublets. We speculate that these predictions might be confounded by LPS induced transcriptional perturbations. Therefore we correct their labels to "Singlets" and retain these cells in the dataset.

```{r fig.width=10, fig.height=7}
p1 <- DimPlot(cells.clean, cells.highlight = WhichCells(cells.clean, expression = doubletfinder == "Doublet")) +
  scale_color_manual(labels = c("NA","Doublets"), values = c("gray","#CC0033")) +
  ggtitle("Predicted Doublets") + NoAxes()
p2 <- DimPlot(cells.clean, group.by = "seurat_clusters", label = T, pt.size = .5) + NoLegend() + NoAxes()
p3 <- FeaturePlot(cells.clean, features = "Cx3cr1") + NoAxes()
p4 <- DimPlot(cells.clean, cells.highlight = WhichCells(cells.clean, expression = seurat_clusters == 12)) + NoLegend() + NoAxes()
(p1 | p2) / (p3 | p4)
```

```{r}
# correct_singlet_ids <- c()
# correct_singlet_ids <- union(correct_singlet_ids, CellSelector(p1))
Idents(cells.clean) <- "doubletfinder"
cells.clean <- SetIdent(cells.clean, cells = correct_singlet_ids, value = "Singlet")
cells.clean$doubletfinder_fixed <- Idents(cells.clean)
cells.filtered <- subset(cells.clean, subset = doubletfinder_fixed != "Doublet")
```

## Quality Control

Using `scater` for QC.

```{r}
Seurat_Obj_Diet <- DietSeurat(cells.filtered)
cells.sce <- as.SingleCellExperiment(Seurat_Obj_Diet)

## Cell level QC
per.cell <- perCellQCMetrics(cells.sce, subsets = list(Mito = grep("^mt-", rownames(cells.sce)),
                                                       Ribo = grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))))
qc.stats <- quickPerCellQC(per.cell, percent_subsets=c("subsets_Mito_percent"))
colSums(as.matrix(qc.stats))

colData(cells.sce) <- cbind(colData(cells.sce), cbind(per.cell, qc.stats))
cells.sce.filtered <- cells.sce[,!cells.sce$discard]

## Feature level QC
keep_feature <- nexprs(cells.sce.filtered, byrow=TRUE) > 0
## Remove mitochondrial and ribosomal genes which does not contribute to analysis (i.e. cell clustering)
keep_feature[grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))] <- FALSE
keep_feature[grep("mt-", rownames(cells.sce))] <- FALSE
cells.sce.filtered <- cells.sce.filtered[keep_feature,]
dim(cells.sce.filtered)

gdko_pbs <- as.Seurat(cells.sce.filtered)
gdko_pbs$ident <- NULL
```

```{r}
gdko_pbs <- gdko_pbs |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Summary plot of some canonical marker genes for typical brain cell types.

```{r}
obj <- gdko_pbs

f_to_plot <- c("nFeature_RNA","Cldn5","Kcnj8","Aqp4","Mbp","Pdgfra",
               "Slc17a7","Slc17a6","Gad2","Cx3cr1","Col1a1",
               "Gsdmd","Casp4","Cd14")
obj <- ScaleData(obj, features = c(VariableFeatures(obj), f_to_plot))
f_list <- FeaturePlot(obj, features = f_to_plot, reduction = "umap", combine = F)
for (i in 1:length(f_list)){
  f_list[[i]] <- f_list[[i]] + 
    NoLegend() +
    NoAxes()
}
f_list <- append(x = f_list, values = list((DimPlot(obj, label = T) + NoLegend() + NoAxes())), after = 0)
```

```{r fig.width=12, fig.height=9}
g <- ggpubr::ggarrange(plotlist = f_list)
g
```

```{r}
ggsave(
  filename = "gdko_pbs_preprocessed_flist.png",
  plot = g,
  device = "png",
  path = "../outs/preprocessing",
  width = 12, height = 10,
  bg = "white"
)
```

## Cell Type Annotation

### Semi-supervised classification

To begin with, we perform an initial round of cell type annotation on the preprocessed, "stand-alone" (e.g. non-integrated) object. These labels will be cross-referenced later in the integrated dataset to check the accuracy of data integration.

```{r}
Idents(gdko_pbs) <- "seurat_clusters"
gdko_pbs <- RenameIdents(
  gdko_pbs, 
  `4` = "Astro", 
  `1` = "Ex-Neu.", `2` = "Ex-Neu.", `3` = "Ex-Neu.", `5` = "Ex-Neu.", `6` = "Ex-Neu.", `7` = "Ex-Neu.", `10` = "Ex-Neu.", `14` = "Ex-Neu.", `18` = "Ex-Neu.", `20` = "Ex-Neu.",
  `11` = "Inh-Neu.", `13` = "Inh-Neu.", `15` = "Inh-Neu.", 
  `12` = "Micro",
  `16` = "BVs", `19` = "BVs",
  `8` = "OPCs",
  `9` = "Oligo", `17` = "Oligo"
  )
gdko_pbs$cell_type_leiden <- Idents(gdko_pbs)
```

```{r fig.width=12, fig.height=5}
d_list <- lapply(levels(gdko_pbs$cell_type_leiden), function(x){
  DimPlot(gdko_pbs, cells.highlight = WhichCells(gdko_pbs, expression = cell_type_leiden == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
ggpubr::ggarrange(plotlist = d_list, ncol = 4, nrow = 2)
```

### Manual curation

Leiden clustering does most of the job just fine.

```{r}
Idents(gdko_pbs) <- "cell_type_leiden"
```

Subsetting out Blood Vessel (BV) population to refine their labels. 

```{r}
## Skip the manual steps and use stashed ids
# bv_ids <- c()
# bv_ids <- union(bv_ids, CellSelector(FeaturePlot(gdko_pbs, features = "Pecam1")))
bv <- subset(gdko_pbs, cells = bv_ids)
DimPlot(gdko_pbs, cells.highlight = bv_ids)
```

```{r fig.width=10, fig.height=5}
bv <- bv |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:20, verbose = FALSE)
```

```{r fig.width=12, fig.height=5}
f_list <- FeaturePlot(bv, features = c("Pecam1","Cldn5","Tnr","Kcnj8","Acta2","Mustn1","Map2","Slc17a7"), combine = FALSE)
ggpubr::ggarrange(plotlist = f_list, ncol = 4, nrow = 2)
```

```{r}
## Skip the manual steps and use stashed ids
# bv_endo_ids <- CellSelector(FeaturePlot(bv, features = "Pecam1"))
# bv_peri_ids <- CellSelector(FeaturePlot(bv, features = "Kcnj8"))
# bv_smc_ids <- CellSelector(FeaturePlot(bv, features = "Acta2"))
# bv_exneu_ids <- CellSelector(FeaturePlot(bv, features = "Slc17a7"))
bv <- SetIdent(bv, cells = bv_endo_ids, value = "Endo")
bv <- SetIdent(bv, cells = bv_peri_ids, value = "Peri")
bv <- SetIdent(bv, cells = bv_smc_ids, value = "SMCs")
bv <- SetIdent(bv, cells = bv_exneu_ids, value = "Ex-Neu.")
bv$cell_type_curated <- Idents(bv)
```

```{r fig.width=9, fig.height=4}
d_list <- lapply(levels(bv$cell_type_curated), function(x){
  DimPlot(bv, cells.highlight = WhichCells(bv, expression = cell_type_curated == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
d_list <- append(x = d_list, values = list((DimPlot(bv) + NoAxes() + NoLegend())), after = 0)
ggpubr::ggarrange(plotlist = d_list, ncol = 3, nrow = 2)
```

```{r}
# Update labels in master object
for(i in levels(bv$cell_type_curated)){
  gdko_pbs <- SetIdent(gdko_pbs, cells = WhichCells(bv, expression = cell_type_curated == i), value = i)
}
```

```{r}
# Stash curated labels
gdko_pbs$cell_type_standalone <- Idents(gdko_pbs)

# Append sample info
gdko_pbs$sample = "GDKO PBS"
gdko_pbs$genotype = "GDKO"
gdko_pbs$treatment = "PBS"
```

## Save Results

```{r}
if(WRITE_CACHE){
  save(correct_singlet_ids, bv_ids, bv_endo_ids, bv_exneu_ids, bv_peri_ids, bv_smc_ids, file = "../miscs/gdko_pbs_ids.rda")
  SeuratDisk::SaveH5Seurat(object = gdko_pbs, filename = "../data/GDKO-PBS-MO-0614_reanalysis/gdko_pbs_preprocessed.h5Seurat", overwrite = T, verbose = F)
}
```

# Session info

```{r}
sessionInfo()
```
