---
title: "Wei et al.: Gsdmd BBB"
subtitle: 'Perturbation Status: Cell Type Prioritization (Augur)'
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'prioritization.html'))})
---

# Introduction

Our experimental setup is based on a two-factor factorial design. To investigate the perturbation strength of LPS challenges across different cell types under wild type or Gsdmd^-/-^ genotypes, we resort to two methods:

  1. Independent non-linear dimension reduction on each cell types using `umap`[^1]^,^[^2]
  2. Cell type prioritization and differential prioritization using `Augur`[^3]

Here, we quantitatively investigate the transcriptional states of each cell type (and their subtypes, if available) within wild-type or Gsdmd^-/-^ mice under the perturbation of LPS, using Augur.

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
WRITE_CACHE = TRUE  ## Set to TRUE on first run
```

```{r}
set.seed(42)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(Augur)
})
```

## Load Data

```{r warning=FALSE}
suppressMessages({
  cells.combined <- SeuratDisk::LoadH5Seurat("../data/Wei_2024/cells_use_combined.h5Seurat", verbose = F)
})
```

# Cell Prioritization

## 1 Prioritize LPS perturbation by "cell_type"

### 1.1 Calculate AUCs for each genotype

```{r}
genotypes = c("WT", "GDKO")
augur_celltype <- sapply(X = genotypes, FUN = function(g){
  # Subset Seurat object
  cells_genotype <- subset(cells.combined, subset = genotype == g)
  
  # Set labels
  cells_genotype$label = cells_genotype$treatment
  
  # Run Augur
  DefaultAssay(cells_genotype) = "RNA"
  augur = calculate_auc(cells_genotype, cell_type_col = "cell_type", n_threads = 32)
  
  return(augur)
  },
  simplify = FALSE, 
  USE.NAMES = TRUE
)
```

### 1.2 Differential Prioritization

#### Generating Null distrubution
```{r}
genotypes = c("WT", "GDKO")
permuted_celltype <- sapply(X = genotypes, FUN = function(g){
  # Subset Seurat object
  cells_pbs <- subset(cells.combined, subset = treatment == "PBS" | genotype == g)
  
  # Set labels
  cells_pbs$label = cells_pbs$treatment
  
  # Run Augur in permutation mode
  augur = calculate_auc(
    cells_pbs, cell_type_col = "cell_type", n_threads = 32, augur_mode = "permute"
  )
  
  return(augur)
  },
  simplify = FALSE, 
  USE.NAMES = TRUE
)
```

#### Comparing observed AUC with permuted results
```{r}
diff_prior_celltype_pvals = calculate_differential_prioritization(
  augur1 = augur_celltype[["WT"]],
  augur2 = augur_celltype[["GDKO"]],
  permuted1 = permuted_celltype[["WT"]],
  permuted2 = permuted_celltype[["GDKO"]]
)
```

## 2 Prioritize LPS perturbation by "cell_subtype"

### 2.1 Calculate AUCs for each genotype

```{r}
# cells.use <- subset(cells.combined, subset = cell_subtype != "Inh-Chodl")  ## Eliminate subtype with small population size
# cells.use$cell_subtype <- droplevels(cells.use$cell_subtype)
genotypes = c("WT", "GDKO")
augur_cellsubtype <- sapply(X = genotypes, FUN = function(g){
  # Subset Seurat object
  cells_genotype <- subset(cells.combined, subset = genotype == g)
  
  # Set labels
  cells_genotype$label = cells_genotype$treatment
  
  # Run Augur
  DefaultAssay(cells_genotype) = "RNA"
  augur = calculate_auc(cells_genotype, cell_type_col = "cell_subtype", n_threads = 32)
  
  return(augur)
  },
  simplify = FALSE, 
  USE.NAMES = TRUE
)
```

### 2.2 Differential Prioritization

#### Generating Null distrubution
```{r}
genotypes = c("WT", "GDKO")
permuted_cellsubtype <- sapply(X = genotypes, FUN = function(g){
  # Subset Seurat object
  cells_null <- subset(cells.combined, subset = treatment == "PBS" | genotype == g)
  
  # Set labels
  cells_null$label = cells_null$treatment
  
  # Run Augur in permutation mode
  augur = calculate_auc(
    cells_null, cell_type_col = "cell_subtype", n_threads = 32, augur_mode = "permute"
  )
  
  return(augur)
  },
  simplify = FALSE, 
  USE.NAMES = TRUE
)
```

#### Comparing observed AUC with permuted results
```{r}
diff_prior_cellsubtype_pvals = calculate_differential_prioritization(
  augur1 = augur_cellsubtype[["WT"]],
  augur2 = augur_cellsubtype[["GDKO"]],
  permuted1 = permuted_cellsubtype[["WT"]],
  permuted2 = permuted_cellsubtype[["GDKO"]]
)
```

## Prioritization of cell subtypes affected by Gsdmd-deficiency under the LPS background

```{r}
cells.lps <- subset(cells.combined, subset = treatment == "LPS")

# Set labels
cells.lps$label = cells.lps$genotype

# Run Augur
DefaultAssay(cells.lps) = "RNA"
augur_cellsubtype_lps_wt_gdko = calculate_auc(cells.lps, cell_type_col = "cell_subtype", n_threads = 32)
```

## Save Results

```{r}
saveRDS(augur_celltype, "../outs/Augur/augur_celltype.rds")
saveRDS(permuted_celltype, "../outs/Augur/permuted_celltype.rds")
saveRDS(augur_cellsubtype, "../outs/Augur/augur_cellsubtype.rds")
saveRDS(permuted_cellsubtype, "../outs/Augur/permuted_cellsubtype.rds")
saveRDS(diff_prior_celltype_pvals, "../outs/Augur/diff_prior_celltype_pvals.rds")
saveRDS(diff_prior_cellsubtype_pvals, "../outs/Augur/diff_prior_cellsubtype_pvals.rds")
saveRDS(augur_cellsubtype_lps_wt_gdko, "../outs/Augur/augur_cellsubtype_lps_wt_gdko.rds")
```


[^1]: McInnes, L., Healy, J. & Melville, J. UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction. arXiv:1802.03426 [cs, stat] (2020).
[^2]: Becht, E. et al. Dimensionality reduction for visualizing single-cell data using UMAP. Nat Biotechnol 37, 38–44 (2019).
[^3]: Skinnider, M. A. et al. Cell type prioritization in single-cell data. Nat Biotechnol (2020) doi:10.1038/s41587-020-0605-1.