---
title: "Wei et al.: Gsdmd BBB"
subtitle: 'Preprocessing workflow for Vanlandewijck (2018)'
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'vanlandewijck_preprocessing.html'))})
---

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
DOWNLOAD_GEO = FALSE  ## Set to TRUE on first run
WRITE_CACHE = TRUE
```

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(scater)
  library(tidyverse)
})
```

```{r}
set.seed(42)
```

# Pre-processing Workflow

We acquire a dataset of brain vascular populations published by Vanlandewijck (2018, Nature). We will download the processed raw count matrix (GSE98816) from GEO.

## Load Data

```{r}
if(DOWNLOAD_GEO){
  suppressPackageStartupMessages({library(GEOquery)})
  options(timeout = 6000)  ## In case of slow Internet connections
  getGEOSuppFiles(GEO = "GSE98816", makeDirectory = T, baseDir = '../data/Vanlandewijck_2018/', filter_regex = "raw")
}
```

```{r}
counts <- read.table(gzfile('../data/Vanlandewijck_2018/GSE98816_Brain_samples_raw_read_counts_matrix.txt.gz'), header = T, row.names = 1)
meta <- data.frame(
  cell_id = colnames(counts),
  tissue = sub("\\..*","",colnames(counts)),
  cell_subtype = sub("_.*","",sub(".*\\.","",colnames(counts)))
  ) |>
  column_to_rownames("cell_id") |>
  mutate(study = "Vanlandewijck")
```

## Standard Seurat Workflow

```{r}
vanlandewijck <- CreateSeuratObject(counts = counts, project = "GD_BBB", meta.data = meta)
```

We don't pay too much attention on the cell subtypes, so we aggregate their labels into primary vessel cell types.

```{r}
Idents(vanlandewijck) <- "cell_subtype"
Idents(vanlandewijck) <- sub(pattern = "[0-9]", replacement = "", Idents(vanlandewijck))
vanlandewijck[["cell_type"]] <- Idents(vanlandewijck)
vanlandewijck[["sample"]] <- "Brain"
```

```{r}
Seurat_Obj_Diet <- DietSeurat(vanlandewijck)
cells.sce <- as.SingleCellExperiment(Seurat_Obj_Diet)

## Cell level QC
per.cell <- perCellQCMetrics(cells.sce, subsets = list(Mito = grep("^mt-", rownames(cells.sce)),
                                                       Ribo = grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))))
qc.stats <- quickPerCellQC(per.cell, percent_subsets=c("subsets_Mito_percent"))
colSums(as.matrix(qc.stats))

colData(cells.sce) <- cbind(colData(cells.sce), cbind(per.cell, qc.stats))
cells.sce.filtered <- cells.sce[,]

## Feature level QC
keep_feature <- nexprs(cells.sce.filtered, byrow=TRUE) > 0
## Remove mitochondrial and ribosomal genes which does not contribute to analysis (i.e. cell clustering)
keep_feature[grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))] <- FALSE
keep_feature[grep("mt-", rownames(cells.sce))] <- FALSE
cells.sce.filtered <- cells.sce.filtered[keep_feature,]
dim(cells.sce.filtered)

vanlandewijck <- as.Seurat(cells.sce.filtered)
vanlandewijck$ident <- NULL
```

```{r}
vanlandewijck <- vanlandewijck |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:30, verbose = FALSE) |>
  RunTSNE(dims = 1:30, verbose = FALSE)
```

```{r}
DimPlot(vanlandewijck, group.by = "cell_type", reduction = "tsne", label = T, label.box = T) + 
  NoAxes() +
  ggtitle(label = "Vanlandewijck et al.: Brain Vascular")
```

## Save Results

```{r}
if(WRITE_CACHE){
  SeuratDisk::SaveH5Seurat(object = vanlandewijck, filename = "../data/Vanlandewijck_2018/vanlandewijck.h5Seurat", overwrite = T, verbose = F)
}
```

# Session info

```{r}
sessionInfo()
```
