---
title: "Wei et al.: Gsdmd BBB"
subtitle: 'Preprocessing workflow for WT-LPS'
author: "Ruiyu Ray Wang"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input_file, encoding){
  out_dir <- '../docs';
  rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'wt_lps_mo_preprocessing.html'))})
---

# Initialization

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
WRITE_CACHE = FALSE  ## Set to TRUE on first run
READ_CACHE = !WRITE_CACHE
```

```{r}
suppressPackageStartupMessages({
  library(SoupX)
  library(Seurat)
  library(DoubletFinder)
  library(scater)
})
```

```{r}
reticulate::use_python("/home/luolab/miniconda3/envs/r-reticulate/bin/python")
```

```{r}
if(!dir.exists("../outs/preprocessing")) {dir.create("../outs/preprocessing")}
```

For knitr to work, steps involving manual operations (i.e. `CellSelector()`) are omitted. Ids for manually selected cells were stashed to and loaded from the `.rda` file located in the `../miscs/` directory.

```{r}
load("../miscs/wt_lps_ids.rda")
```

```{r}
set.seed(42)
```

# Pre-processing Workflow

## Decontamination

Using `SoupX` for decontamination. Before constructing the soupchannel, precompute Louvain clusters and umap embeddings which will be later used.

```{r}
filtered_counts.wt_lps <- Read10X("../data/WT-LPS-MO-0614/outs/filtered_feature_bc_matrix/")
cells <- CreateSeuratObject(counts = filtered_counts.wt_lps, project = "GD_BBB")
# Standard Seurat Workflow. For conciseness, future runs of the workflow will use verbose = FALSE
cells <- cells |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA() |>
  FindNeighbors(dims = 1:50) |>
  FindClusters() |>
  RunUMAP(dims = 1:50)
```

```{r}
# Load data and estimate soup profile
sc = load10X("../data/WT-LPS-MO-0614/outs/")
# Set clusters and DR
sc = setClusters(sc, cells$RNA_snn_res.0.8)
sc = setDR(sc, Embeddings(object = cells, reduction = "umap")[colnames(sc$toc), c("UMAP_1","UMAP_2")])
# Estimate rho
sc = autoEstCont(sc)
# Clean the data
out = adjustCounts(sc, roundToInt = T)  # stochastic, don't forget to set.seed() for reproducibility
```

```{r}
if(WRITE_CACHE){
  DropletUtils::write10xCounts(path = "../data/WT-LPS-MO-0614/outs/strainedCounts.h5", x = out, type = "HDF5", overwrite = TRUE)  ## Save corrected matrix
}
```

```{r}
if(READ_CACHE){out <- Read10X_h5("../data/WT-LPS-MO-0614/outs/strainedCounts.h5")}

cells.decont <- CreateSeuratObject(counts = out, project = "GD_BBB")
cells.decont <- cells.decont |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Here, we're dealing with a sample under acute LPS treatment. This will cause the cell to undergo apoptosis and mRNA leakage. Unsuprisingly, we find many putatively 'Soup' droplets from our data. By iterative sub-clustering, we confirm that they correspond to cluster 1,3,4,8,13,26 and 27. These clusters display low level of recovered genes (nFeature_RNA) and high level of mitochondrial RNA species. Here, we filter them out manually.

```{r fig.width=9, fig.height=6}
((DimPlot(cells.decont, group.by = "seurat_clusters", label = T) + NoLegend()) | DimPlot(cells.decont, cells.highlight = WhichCells(cells.decont, idents = c(1,3,4,8,13,26,27))) + ggtitle("Cells to Remove\n(Putative Empty droplets)") + NoLegend())/
  (FeaturePlot(cells.decont, features = c("nFeature_RNA","mt-Nd3")))
```

```{r}
cells.clean <- subset(
  cells.decont,
  idents = c(1,3,4,8,13,26,27),
  invert = TRUE
)
```

## Doublet Removal

We use `DoubletFinder` to remove doublets. We aimed at ~8000 cell recovery during wet experiment. According to 10X, this corresponds to ~0.06 doublet rates, thus we set `percent_est = 0.06`.

```{r}
PCs = 1:10
percent_est = 0.06
pN = 0.25
```

```{r}
cells.clean <- cells.clean |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

```{r}
suppressed_msg <- capture.output({
  sweep.res.list_cells <- paramSweep_v3(cells.clean, PCs = PCs, sct = FALSE)
  sweep.stats_cell <- summarizeSweep(sweep.res.list_cells, GT = FALSE)
  bcmvn_cell <- find.pK(sweep.stats_cell)
  pk_cell <- as.numeric(levels(bcmvn_cell[, 2]))[which.max(bcmvn_cell[,5])]
  homotypic.prop <- modelHomotypic(annotations <- cells.clean@meta.data$seurat_clusters)
  nExp_poi <- round(percent_est*nrow(cells.clean@meta.data))
  nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
  cells.clean <- doubletFinder_v3(cells.clean, PCs, pN, pK = pk_cell, nExp = nExp_poi.adj, reuse.pANN = FALSE, sct = FALSE)
  cells.clean$doubletfinder <- cells.clean@meta.data[grep("DF.",names(cells.clean@meta.data))]
  # cells.filtered <- subset(cells.clean, subset = doubletfinder != "Doublet")
})
```

Beware, sometimes we may encounter over removal! As a visual check, inspect the predicted doublets. We notice that two large clusters corresponding to Microglia/Macrophage (2) and Blood Vessel Cells (12) contain many cells predicted as doublets. We speculate that these predictions might be confounded by LPS induced transcriptional perturbations. Therefore we correct their labels to "Singlets" and retain these cells in the dataset.

```{r}
p1 <- DimPlot(cells.clean, cells.highlight = WhichCells(cells.clean, expression = doubletfinder == "Doublet")) +
  scale_color_manual(labels = c("NA","Doublets"), values = c("gray","#CC0033")) +
  ggtitle("Predicted Doublets") + NoAxes()
p2 <- DimPlot(cells.clean, group.by = "seurat_clusters", label = T) + NoLegend() + NoAxes()
p3 <- FeaturePlot(cells.clean, features = "Cx3cr1") + NoAxes()
p4 <- FeaturePlot(cells.clean, features = "Pecam1") + NoAxes()
(p1 | p2) / (p3 | p4)
```

```{r}
# correct_singlet_ids <- CellSelector(p1)
Idents(cells.clean) <- "doubletfinder"
cells.clean <- SetIdent(cells.clean, cells = correct_singlet_ids, value = "Singlet")
cells.clean$doubletfinder_fixed <- Idents(cells.clean)
cells.filtered <- subset(cells.clean, subset = doubletfinder_fixed != "Doublet")
```

## Quality Control

Using `scater` for QC.

```{r}
Seurat_Obj_Diet <- DietSeurat(cells.filtered)
cells.sce <- as.SingleCellExperiment(Seurat_Obj_Diet)

## Cell level QC
per.cell <- perCellQCMetrics(cells.sce, subsets = list(Mito = grep("^mt-", rownames(cells.sce)),
                                                       Ribo = grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))))
qc.stats <- quickPerCellQC(per.cell, percent_subsets=c("subsets_Mito_percent"))
colSums(as.matrix(qc.stats))

colData(cells.sce) <- cbind(colData(cells.sce), cbind(per.cell, qc.stats))
cells.sce.filtered <- cells.sce[,!cells.sce$discard]

## Feature level QC
keep_feature <- nexprs(cells.sce.filtered, byrow=TRUE) > 0
## Remove mitochondrial and ribosomal genes which does not contribute to analysis (i.e. cell clustering)
keep_feature[grep("\\bRp[sl]\\d+[^k]*\\b", rownames(cells.sce))] <- FALSE
keep_feature[grep("mt-", rownames(cells.sce))] <- FALSE
cells.sce.filtered <- cells.sce.filtered[keep_feature,]
dim(cells.sce.filtered)

wt_lps <- as.Seurat(cells.sce.filtered)
wt_lps$ident <- NULL
```

```{r}
wt_lps <- wt_lps |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:50, verbose = FALSE) |>
  FindClusters(algorithm = 4, verbose = FALSE) |>
  RunUMAP(dims = 1:50, verbose = FALSE)
```

Summary plot of some canonical marker genes for typical brain cell types.

```{r}
obj <- wt_lps

f_to_plot <- c("nFeature_RNA","Cldn5","Kcnj8","Aqp4","Mbp","Pdgfra",
               "Slc17a7","Slc17a6","Gad2","Cx3cr1","Col1a1",
               "Gsdmd","Casp4","Cd14")
obj <- ScaleData(obj, features = c(VariableFeatures(obj), f_to_plot))
f_list <- FeaturePlot(obj, features = f_to_plot, reduction = "umap", combine = F)
for (i in 1:length(f_list)){
  f_list[[i]] <- f_list[[i]] + 
    NoLegend() +
    NoAxes()
}
f_list <- append(x = f_list, values = list((DimPlot(obj, label = T) + NoLegend() + NoAxes())), after = 0)
```

```{r fig.width=12, fig.height=10}
g <- ggpubr::ggarrange(plotlist = f_list)
g
```

```{r}
ggsave(
  filename = "wt_lps_preprocessed_flist.png",
  plot = g,
  device = "png",
  path = "../outs/preprocessing",
  width = 12, height = 10,
  bg = "white"
)
```

## Cell Type Annotation

### Semi-supervised classification

To begin with, we perform an initial round of cell type annotation on the preprocessed, "stand-alone" (e.g. non-integrated) object. These labels will be cross-referenced later in the integrated dataset to check the accuracy of data integration.

```{r}
Idents(wt_lps) <- "seurat_clusters"
wt_lps <- RenameIdents(
  wt_lps, 
  `14` = "Astro", 
  `1` = "Ex-Neu.", `3` = "Ex-Neu.", `5` = "Ex-Neu.", `7` = "Ex-Neu.", `9` = "Ex-Neu.", `16` = "Ex-Neu.", `17` = "Ex-Neu.", `21` = "Ex-Neu.", `24` = "Ex-Neu.",
  `6` = "Inh-Neu.", `10` = "Inh-Neu.", `11` = "Inh-Neu.", `13` = "Inh-Neu.", `15` = "Inh-Neu.", `18` = "Inh-Neu.", `22` = "Inh-Neu.", `23` = "Inh-Neu.",
  `4` = "Micro",
  `12` = "BVs",
  `8` = "OPCs",
  `2` = "Oligo", `19` = "Oligo", `20` = "Oligo"
  )
wt_lps$cell_type_leiden <- Idents(wt_lps)
```

```{r fig.width=12, fig.height=5}
d_list <- lapply(levels(wt_lps$cell_type_leiden), function(x){
  DimPlot(wt_lps, cells.highlight = WhichCells(wt_lps, expression = cell_type_leiden == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
ggpubr::ggarrange(plotlist = d_list, ncol = 4, nrow = 2)
```

### Manual curation

Some cell appear to be mis-allocated with respect to their identity. To fix this, we perform iterative sub-clustering and manually curate the labels of these cells.

```{r}
Idents(wt_lps) <- "cell_type_leiden"
```

For blood vessel associated populations, the `BVs` cluster can be further divided into Endothelial cells (Endo), Pericytes (Peri), Smooth Muscle Cells (SMCs) and Vascular Leptomeningeal Cells (VLMCs), with respect to their markers (shown below). Additionally, some Oligodendrocytes (Mbp-positive) and Neurons (Slc17a7 or Gad2-positive) was mis-placed into this cluster by umap. We also check that their labels here.

```{r}
## Skip the manual steps and use stashed ids
# bv_ids <- CellSelector(FeaturePlot(wt_lps, features = "Pecam1"))
DimPlot(wt_lps, cells.highlight = bv_ids)
# bv <- subset(wt_lps, cells = CellSelector(DimPlot(wt_lps)))
bv <- subset(wt_lps, cells = bv_ids)
bv <- bv |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:20, verbose = FALSE)
```

```{r fig.width=12, fig.height=7.5}
f_list <- FeaturePlot(bv, features = c("Pecam1","Cldn5","Tnr","Kcnj8","Acta2","Mustn1","Col1a1","Slc6a13","Gad2","Slc17a7","Map2","Mbp"), combine = FALSE)
ggpubr::ggarrange(plotlist = f_list, ncol = 4, nrow = 3)
```

```{r}
## Skip the manual steps and use stashed ids
# endo_ids <- c()
# endo_ids <- union(endo_ids, CellSelector(FeaturePlot(bv, features = "Pecam1")))
# peri_ids <- CellSelector(FeaturePlot(bv, features = "Kcnj8"))
# vlmc_ids <- CellSelector(FeaturePlot(bv, features = "Col1a1"))
# smc_ids <- CellSelector(FeaturePlot(bv, features = "Acta2"))
# oligo_ids <- CellSelector(FeaturePlot(bv, features = "Mbp"))
# ex_neu_ids <- c()
# ex_neu_ids <- union(ex_neu_ids, CellSelector(FeaturePlot(bv, features = "Slc17a7")))
# inh_neu_ids <- CellSelector(FeaturePlot(bv, features = "Gad2"))
bv <- SetIdent(bv, cells = bv_endo_ids, value = "Endo")
bv <- SetIdent(bv, cells = bv_peri_ids, value = "Peri")
bv <- SetIdent(bv, cells = bv_vlmc_ids, value = "VLMCs")
bv <- SetIdent(bv, cells = bv_smc_ids, value = "SMCs")
bv <- SetIdent(bv, cells = bv_oligo_ids, value = "Oligo")
bv <- SetIdent(bv, cells = bv_ex_neu_ids, value = "Ex-Neu.")
bv <- SetIdent(bv, cells = bv_inh_neu_ids, value = "Inh-Neu.")
bv$cell_type_curated <- Idents(bv)
```

```{r fig.width=12, fig.height=5}
d_list <- lapply(levels(bv$cell_type_curated), function(x){
  DimPlot(bv, cells.highlight = WhichCells(bv, expression = cell_type_curated == x)) + NoLegend() + NoAxes() + ggtitle(x)
})
d_list <- append(x = d_list, values = list((DimPlot(bv) + NoAxes() + NoLegend())), after = 0)
ggpubr::ggarrange(plotlist = d_list, ncol = 4, nrow = 2)
```

```{r}
# Update labels in master object
for(i in levels(bv$cell_type_curated)){
  wt_lps <- SetIdent(wt_lps, cells = WhichCells(bv, expression = cell_type_curated == i), value = i)
}
```

For Microglia/Macrophage, we refer to them as the Monocyte population and refine their labels with respect to a canonical marker, Mrc1, that readily distinguishes the Microglia and PVMs (Perivascular Macrophage).

```{r}
## Skip the manual steps and use stashed ids
# mono_ids <- CellSelector(FeaturePlot(wt_lps, features = "Cx3cr1"))
DimPlot(wt_lps, cells.highlight = mono_ids)
mono <- subset(wt_lps, cells = mono_ids)
mono <- mono |>
  NormalizeData(verbose = FALSE) |>
  FindVariableFeatures(verbose = FALSE) |>
  ScaleData(verbose = FALSE) |>
  RunPCA(verbose = FALSE) |>
  FindNeighbors(dims = 1:20, verbose = FALSE) |>
  FindClusters(verbose = FALSE) |>
  RunUMAP(dims = 1:20, verbose = FALSE)
```

```{r}
# mono_pvm_ids <- CellSelector(FeaturePlot(mono, features = "Mrc1"))
Idents(mono) <- "cell_type_leiden"
mono <- SetIdent(mono, cells = mono_pvm_ids, value = "PVMs")
mono$cell_type_curated <- Idents(mono)
```

```{r}
# Update labels in master object
for(i in levels(mono$cell_type_curated)){
  wt_lps <- SetIdent(wt_lps, cells = WhichCells(mono, expression = cell_type_curated == i), value = i)
}
```

```{r}
# Stash curated labels
wt_lps$cell_type_standalone <- Idents(wt_lps)

# Append sample info
wt_lps$sample = "WT LPS"
wt_lps$genotype = "WT"
wt_lps$treatment = "LPS"
```

## Save Results

```{r}
if(WRITE_CACHE){
  save(correct_singlet_ids, bv_ids, bv_endo_ids, bv_ex_neu_ids, bv_inh_neu_ids, bv_oligo_ids, bv_peri_ids, bv_smc_ids, bv_vlmc_ids, mono_ids, mono_pvm_ids, 
       file = "../miscs/wt_lps_ids.rda")
  SeuratDisk::SaveH5Seurat(object = wt_lps, filename = "../data/WT-LPS-MO-0614/wt_lps_preprocessed.h5Seurat", overwrite = T, verbose = F)
}
```

# Session info

```{r}
sessionInfo()
```
